Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-04-03 paper.Rnw
Small updates to the methods wording, Park QC added, bad a2a_tracc data added

@@

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT}
\author{Pipitone J., Winterburn J., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

@@

Our algorith, called MAGeT brain ({\em M}ultiple {\em A}utomatically {\em
Ge}nerated {\em T}emplates), is an extension to the basic multi-atlas-based
segmentation schema\citep{Chakravarty2012}. Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the targets in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion. The
intuition driving this approach is that by generating a template library we
leverage the unique neuroanatomy of target population on hand to initialize the
segmentation process and improve accuracy over direct propagation from the atlas
library to unlabelled targets while also using fewer manually segmented
atlases.

@@

In previous work \citep{Chakravarty2012}, we applied MAGeT brain to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
MAGeT and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of MAGeT brain under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate MAGeT
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five in-vivo high-resolution (300u
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate MAGeT in a
real world situation, we segment the entire ADNI1 screening dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure MAGeT brain accuracy across disease categories, we also
compare MAGeT segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsection{MAGeT Brain Algorithm}
In this paper we will be making a distinction between an atlas and a template --
typically these terms are used roughly interchangeably. The term {\it atlas} is
taken to refer to two image volumes: an intensity image ({\it atlas image}) and
a corresponding manual segmentation image ({\it atlas labels}). {\it Template}
refers more generically to any image and corresponding labelling, manual or
computed, when it is {\it used as} as a model in the segmentation of another
image.  The terms {\it atlas library} and {\it template library} mean a set of
such images. Additionally, we will use the terms {\it target} to refer to an
intensity image for which we would like an segmentation.

@@

MAGeT brain is best understood as an extension of the basic multi-atlas
segmentation schema. Instead of using the atlas library to directly label the
target images, a subset of the input images are selected as template images and
then labelled.  The choice of targets used in the template library can be made
to reflect the neuroanatomy or demographics of the target set as a whole (for
instance, by sampling equally from cases and controls). Once the template
library images have been chosen, a truncated version of basic multi-atlas
segmentation is used to label the template library images without performing
label fusion. Instead, each template image receives multiple labellings: one
from each atlas image. A second round of basic multi-atlas segmentation uses the
template library to segment the entire set of target images (including those
images used in the template library). Label fusion in this final step fuses all
labels from all templates. To summarize, figure \ref{alg:MAGeT} describes the
MAGeT brain algorithm in pseudocode.

Source code for MAGeT brain can be found at
\url{http://github.com/pipitone/MAGeTbrain}.

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the MAGeT Brain algorithm}
  \label{alg:MAGeT}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$target$}
        \ForAll{$template$}
          \State propagate all labels for template to target space
          \State store target labels
        \EndFor
        \State fuse target labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a target to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

The following experiments were performed to assess the performance of MAGeT
brain with various parameter settings as well as on diverse datasets.  In each
experiment we contrast the performance of MAGeT brain with that standard single-
and multi-atlas segmentations derived from the same atlas library.

%%%%%
%%%%%                 Experiments: ADNI-1 cross-validationn               %%%%%
%%%%%
\subsubsection{ADNI1 cross-validation}
To test the effect of parameter settings on the MAGeT brain algorithm, Monte
Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a subset of
the ADNI1 subjects for validation.  This form of cross-validation (described in
detail below) allows us to validate MAGeT with various atlas and template
library sizes, registration and label fusion methods.

@@

{\bf Atlas and template libary.}  Atlases consist of images taken from the
dataset, with corresponding manual labels provided by SNT.  Atlas library size
varied from 3 to 9 images.  The remaining 60-66 target images were segmented
with the template library size varying from 3 to 20 images for each setting of
atlas library size. Template library images were selected at random from the
targets.

@@

{\bf Label fusion.} Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote were used. With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 3 to 20 images.

{\bf Evaluation.}  Monte Carlo Cross-Validation (MCCV), also known as repeated
random sub-sampling cross-validation, consists of repeated trials in which items
from the dataset are randomly sampled (without replacement) and assigned to a
training set or validation set \cite{Shao1993}.

In each trial, the training set consists of images and SNT labels used as
atlases, and the validation set consists of images to be segmented along with
the corresponding SNT labels used as gold-standard segmentations. Both MAGeT
brain and basic multi-atlas segmentation were performed in each trial, and the
accuracy of the resulting segmentations were measured against the SNT labels.

We performed 10 trials on each combination of parameters listed above: atlas
library size (1-9), template library size (1-20), registration method (ANTS vs.
ANIMAL), and label fusion method (MV, XC-WV, NMI-WV). A total of 1080 parameter
settings were evaluated over a total of 10800 trials.

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
To test the accuracy of MAGeT brain on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

@@

{\bf Evaluation.}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct similarity comparisons
between labels produced by MAGeT brain and the ADNI labels.  To evaluate the
performance of MAGeT brain, we compare the correlation of MAGeT segmentation
volumes with manual segmentation (SNT) volumes. Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to MAGeT
segmentations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI1-scr-tmpllib-demographics, echo=F, results="asis",cache=FALSE>>=
scr_tmpllib_RIDs = c(0295, 0413, 0619, 0685, 0729, 0782, 0938, 0954, 1018, 1155,
                 0907, 0981, 0448, 0553, 0814, 1130, 0698, 1339, 1095, 0991,
                 0159)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               scr_tmpllib_RIDs, method = "reverse", test=FALSE, overall=TRUE)

@@

{\bf Dataset evaluated.} To validate that MAGeT performance generalises to
across disease categories, we measure the performance of MAGeT on the
Schizophrenic First Episode Patient (SFEP) dataset.

@@

{\bf Evaluation.}  - Leave one out cross-validation (LOOCV) in which all five
subjects are segmented in separate runs of MAGeT brain.  In each run, the subject
to be segmented is excluded from the Atlas library (so only four atlases are used).

@@

%                                                                              %
% ADNI1 Screening Validation - results preparation                             %
%                                                                              %
<<ADN1-Xval-prep, include=FALSE, cache=TRUE>>=
# all data from the validation trials
t_data       = read.csv(gzfile('data/a2a_ants/results_2013_01_03.csv.gz'))
t_data       = subset(t_data, atlases %in% c(1,3,5,7,9))
a_data       = t_data[,] #read.csv(gzfile('data/a2a_tracc/results_2013_01_03.csv.gz'))
diagnoses    = read.csv('data/a2a_diagnoses.csv')

# simplifies the table
t_data$se    = NULL
t_data$sn    = NULL
t_data$j     = NULL
a_data$se    = NULL
a_data$sn    = NULL
a_data$j     = NULL

@@

# separate MA majvote from MB so we can compute differences
t_ma = subset(t_data, approach=="multiatlas" & method=="majvote")
a_ma = subset(a_data, approach=="multiatlas" & method=="majvote")
t_mb = subset(t_data, approach=="mb")
a_mb = subset(a_data, approach=="mb")

@@

all_data = merge(ma, mb, by=c("timestamp", "atlases", "batch", "label", "reg_method", "subject"))
all_data$k_diff = all_data$k.y - all_data$k.x 
method_labels <- list("Majority Vote" = "majvote", 
                      "NMI Vote" = "nmi", 
                      "Cross-correlation Vote" = "xcorr")
levels(all_data$method.y) <- method_labels
all_data       = merge(all_data, diagnoses)
remove(t_ma, a_ma, t_mb, a_mb, ma, mb)
@
%                                                                              %
%                                                                              %

% TODO: Bland-Altman plot for MA and MAGeT

Hippocampal MAGeT-based segmentations using both ANIMAL and ANTs registration
algorithm demonstrate good overlap with manually derived gold-standards ( Figure
\ref{fig:ADNI1-xval-k-mean}, \todo{table?}) using each label fusion method.
Qualitatively, both ANIMAL and ANTs-based segmentations demonstrate trend
overlap accuracy that increases with the size of atlas library and template
library, with at least 12 templates needed for optimal results
\todo{quantify?}).  No significant difference in segmentation accuracy is seen
when either ANIMAL or ANTs registration is used with different label fusion
techniques, at any atlas or template library sizes (\todo{lm of interaction?).
In every parameter configuration, the use of MAGeT Brain with ANTs registration
shows increased segmentation accuracy over MAGeT Brain with ANIMAL registration
(\todo{interaction?}).

It is interesting to note that with an even number of templates, MAGeT brain
shows a small decrease in performance relative to when one fewer template image
is used (\todo{show significance of this}). See section (\todo{ref}) for a
discussion of this behaviour. In the remainder of this section, only results
from odd-sized template libraries will be shown.

@@

ggplot(subset(all_data, method == 'majvote'), 
       aes(x=templates.y,y=k.y, colour=as.factor(atlases))) + 
  facet_grid(reg_method~method.y) + 
  #stat_summary(fun.y=mean,geom='line',
  #             aes(y=k.y)) + 
  stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" )
@
  \caption{MAGeT brain mean DSC by atlas and template library size}
  \label{fig:ADNI1_xval-k_mean}
\end{figure}

With an increasing number of templates, MAGeT-based segmentating using ANTs
registration and majority vote label fusion shows improvement in overlap
accuracy over multi-atlas-based segmentation, using the same number of atlases
and voting method Figure \todo{ref}. The magnitude of improvement over
multi-atlas-based segmentation decreases with an increasing number of atlases,
with accuracy converging with 7 atlases.  Peak improvement in MAGeT Brain
accuracy (0.02 DSC) is found when one atlas is used with a template library of
20 images. \todo{stats?}

\begin{figure}
<<ADN1-Xval-k-diff, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ggplot(subset(all_data, reg_method=="ANTS" & method.y=="Majority Vote"), 
       aes(x=templates.y, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  coord_cartesian(ylim=c(-0.1,0.05)) + 
  scale_x_continuous(breaks=1:20) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(legend=FALSE) + 
  scale_linetype_discrete(legend=FALSE) + 
  xlab( "Number of Templates" ) + 
  ylab( "Increase in mean Kappa between MAGeT brain and multiatlas" ) 
@
  \caption{Increase in mean Kappa between MAGeT brain and multi-atlas}
  \label{ADNI1_cross_validation_MA_difference}
\end{figure}

In addition to an improvement in accuracy over multi-atlas-based segmentation,
MAGeT Brain also shows a decrease in the variability of segmentation accuracy
with different input atlases (Figure \todo{ref}).  Additionally, variability
decreases with the size of template library, and is significantly decreased
(p<0.05) when using 3-7 atlases and 17 or more templates.

\begin{figure}
<<ADN1-Xval-k-sd-t-test-p-plot, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
    ANTSmajvote = subset(all_data, reg_method=="ANTS" & method.y=="Majority Vote")

stddev = ddply(ANTSmajvote, c("subject", "label","atlases","templates.y"), function (df) {
  data.frame(
    MA = sd(df$k.x),
    MB = sd(df$k.y) 
  )
})

stddev.t.test = ddply(stddev, c("atlases","templates.y"), function (df) {
  t = t.test(df$MA, df$MB)
  data.frame(
    statistic = t$statistic, 
    p.value = t$p.value
  )
})

ggplot(stddev.t.test, 
       aes(x=templates.y, y=p.value, colour=as.factor(atlases))) + 
  geom_line() + 
  geom_hline(aes(yintercept=0.05), linetype='dashed') + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_x_continuous(breaks=1:20) + scale_y_continuous(breaks=seq(0,1,by=0.05)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) 
@
\end{figure}

@@

\begin{figure}
<<ADN1-Xval-k-diff-ANTS-20tmpls, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.y    == "Majority Vote" &
                      templates.y == 20)
ggplot(a_20_tmpls, aes(x=factor(atlases))) +
  stat_summary(fun.y=mean,geom='point',size=3,colour="red",aes(y=k.x)) +
  stat_summary(fun.y=mean,geom='line',colour="red",aes(y=k.x,group=1)) +
  stat_summary(fun.y=mean,geom='point',size=3,colour="blue",aes(y=k.y)) +
  stat_summary(fun.y=mean,geom='line',colour="blue",aes(y=k.y,group=1)) +
  labs(x="Number of Atlases", y = "Kappa", 
       title="mean Kappa of MA and MAGeT brain (ANTS/20 templates) \nvs\natlas library size") +
  scale_colour_hue(name="Method") + 
  coord_cartesian(ylim=c(0.7,0.9))
  #theme(legend.justification=c(1,0), legend.position=c(1,0))
@
  \caption{Change in mean Kappa between multi-atlas and MAGeT brain using ANTS,
  20 templates}
  \label{}
\end{figure}

>>>>

\newcommand{\mb}{MAGeT Brain}

@@

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author{Pipitone J., Winterburn J., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

@@

Our algorith, called \mb ({\em M}ultiple {\em A}utomatically {\em
Ge}nerated {\em T}emplates), is an extension to the basic multi-atlas-based
segmentation schema\citep{Chakravarty2012}. Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the targets in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion. The
intuition driving this approach is that by generating a template library we
leverage the unique neuroanatomy of target population on hand to initialize the
segmentation process and improve accuracy over direct propagation from the atlas
library to unlabelled targets while also using fewer manually segmented
atlases.

@@

In previous work \citep{Chakravarty2012}, we applied \mb to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
\mb and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of \mb under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate \mb
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five in-vivo high-resolution (300u
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate \mb in a
real world situation, we segment the entire ADNI1 screening dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure \mb accuracy across disease categories, we also
compare \mb segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsection{\mb Algorithm}
In this paper we will be making a distinction between an atlas and a template --
typically these terms are used roughly interchangeably. The term {\it atlas} is
taken to refer to two image volumes: an intensity image ({\it atlas image}) and
a corresponding manual segmentation image ({\it atlas labels}). {\it Template}
refers more generically to any image and corresponding labelling, manual or
computed, when it is {\it used as} as a model in the segmentation of another
image.  The terms {\it atlas library} and {\it template library} mean a set of
such images. Additionally, we will use the terms {\it target} to refer to an
intensity image for which we would like an segmentation.

@@

\mb is best understood as an extension of the basic multi-atlas
segmentation schema. Instead of using the atlas library to directly label the
target images, a subset of the input images are selected as template images and
then labelled.  The choice of targets used in the template library can be made
to reflect the neuroanatomy or demographics of the target set as a whole (for
instance, by sampling equally from cases and controls). Once the template
library images have been chosen, a truncated version of basic multi-atlas
segmentation is used to label the template library images without performing
label fusion. Instead, each template image receives multiple labellings: one
from each atlas image. A second round of basic multi-atlas segmentation uses the
template library to segment the entire set of target images (including those
images used in the template library). Label fusion in this final step fuses all
labels from all templates. To summarize, figure \ref{alg:MAGeT} describes the
\mb algorithm in pseudocode.

Source code for \mb can be found at
\url{http://github.com/pipitone/MAGeTbrain}.

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the \mb algorithm}
  \label{alg:MAGeT}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$target$}
        \ForAll{$template$}
          \State propagate all labels for template to target space
          \State store target labels
        \EndFor
        \State fuse target labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a target to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

The following experiments were performed to assess the performance of \mb with
various parameter settings as well as on diverse datasets.  In each experiment
we contrast the performance of \mb with that standard single- and multi-atlas
segmentations derived from the same atlas library.

%%%%%
%%%%%                 Experiments: ADNI-1 cross-validationn               %%%%%
%%%%%
\subsubsection{ADNI1 cross-validation}
To test the effect of parameter settings on the \mb algorithm, Monte
Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a subset of
the ADNI1 subjects for validation.  This form of cross-validation (described in
detail below) allows us to validate \mb with various atlas and template
library sizes, registration and label fusion methods.

@@

{\bf Atlas and template libary.}  The atlas and template libraries are selected
from the dataset evaluated.  Atlases consist of images taken from the
dataset, with corresponding manual labels provided by SNT. Template library 
images were selected at random from remaining images not used as atlases. The
atlas library size ranged from 1 to 9 images. Template library size ranged from
1 to 20 images.

@@

{\bf Label fusion.} Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote were used. With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 1 to 20 images.

{\bf Evaluation.}  Monte Carlo Cross-Validation (MCCV), also known as repeated
random sub-sampling cross-validation, consists of repeated rounds of validation 
in which items from the dataset are randomly sampled (without replacement) and 
assigned to a training set or validation set \cite{Shao1993}.

In this experiment, each validation round consists of the following steps. First, 
an atlas library is selected (the training set) and, from the remaining images, 
a template library and subject to be segmented is chosen (the validation set). 
Second, \mb and basic multi-atlas segmentation are performed on the subject using the 
selected atlas and template libraries. Third, the accuracy of the resulting 
segmentations are measured against the SNT labels.

A total of 10 validation rounds are perfomed on each subject in the dataset, for each 
combination of parameter settings: atlas library size (1-9), template library 
size (1-20), registration method (ANTS vs. ANIMAL), and label fusion method 
(MV, XC-WV, NMI-WV). A total of $10*69*9*20*2*3 = 7452000$ validation rounds were
conducted, resulting in a total of 1'490'400 segmentations analysed.

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
To test the accuracy of \mb on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

@@

{\bf Evaluation.}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct similarity comparisons
between labels produced by \mb and the ADNI labels.  To evaluate the
performance of \mb, we compare the correlation of \mb segmentation
volumes with manual segmentation (SNT) volumes. Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to \mb
segmentations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Template Library Demographics Table
%
<<ADNI1-scr-tmpllib-demographics, echo=F, results="asis",cache=FALSE>>=
scr_tmpllib_RIDs = c(0295, 0413, 0619, 0685, 0729, 0782, 0938, 0954, 1018, 1155,
                 0907, 0981, 0448, 0553, 0814, 1130, 0698, 1339, 1095, 0991,
                 0159)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               scr_tmpllib_RIDs, method = "reverse", test=FALSE, overall=TRUE)

@@

{\bf Dataset evaluated.} To validate that \mb performance generalises to
across disease categories, we measure the performance of \mb on the
Schizophrenic First Episode Patient (SFEP) dataset.

@@

{\bf Evaluation.}  - Leave one out cross-validation (LOOCV) in which all five
subjects are segmented in separate runs of \mb.  In each run, the subject
to be segmented is excluded from the Atlas library (so only four atlases are used).

@@

%                                                                              %
% ADNI1 Screening Validation - results preparation                             %
%                                                                              %
<<ADN1-xval-prep, include=FALSE, cache=TRUE>>=
# all data from the validation trials
a_data       = read.csv(gzfile('data/a2a_ants/results_2013_01_03.csv.gz'))
#t_data       = read.csv(gzfile('data/a2a_tracc/xfmtrials_results_2013-04-01.csv.gz'))
t_data       = read.csv(gzfile('data/a2a_tracc/results_2012_12_31.csv.gz'))
diagnoses    = read.csv('data/a2a_diagnoses.csv')

# simplifies the table
t_data       = subset(t_data, atlases %in% c(1,3,5,7,9))
a_data       = subset(a_data, atlases %in% c(1,3,5,7,9))
t_data$se    = NULL
t_data$sn    = NULL
t_data$j     = NULL
a_data$se    = NULL
a_data$sn    = NULL
a_data$j     = NULL

@@

# separate MA majvote from MB so we can compute differences
# we only compare multiatlas using majority vote
t_ma = subset(t_data, approach=="multiatlas" & method=="majvote")
a_ma = subset(a_data, approach=="multiatlas" & method=="majvote")
t_mb = subset(t_data, approach=="mb")
a_mb = subset(a_data, approach=="mb")

@@

# aggregate over batches
ma_mean = aggregate(k ~ subject + reg_method + approach + method + atlases + templates + top_n + label, data = ma, mean)
mb_mean = aggregate(k ~ subject + reg_method + approach + method + atlases + templates + top_n + label, data = mb, mean)

all_data = merge(ma, mb, by=c("timestamp", "atlases", "batch", "label", "reg_method", "subject"), 
                 suffixes=c(".ma", ".mb"))

all_data_mean = merge(ma_mean, mb_mean, by=c("reg_method", "label", "atlases", "subject"), 
                 suffixes=c(".ma", ".mb"))

all_data$k_diff = all_data$k.mb - all_data$k.ma
all_data_mean$k_diff = all_data_mean$k.mb - all_data_mean$k.ma

vote_levels <- list("Majority Vote" = "majvote", 
                      "NMI Vote" = "nmi", 
                      "Cross-correlation Vote" = "xcorr")
levels(all_data$method.mb) <- vote_levels
levels(all_data_mean$method.mb) <- vote_levels

all_data       = merge(all_data, diagnoses)
all_data_mean  = merge(all_data_mean, diagnoses)
#remove(t_ma, a_ma, t_mb, a_mb, ma, mb)
@
%                                                                              %
%                                                                              %

%
% synonyms for significant: 
% powerful, suggestive, apparent, definite, evident, large, marked, material, 
% noticeable, plain, pronounced, substantial, substantive, 
%
% antonyms: 
% negligible, minor, no appreciable, scant

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of 
each of 69 subjects, over a range of atlas and template library sizes, with ANTS 
and ANIMAL registration algorithms, and three kinds of label fusion techniques. 
Hippocampal \mb-based segmentations using both ANIMAL and ANTs registration
algorithm demonstrate good overlap with manually derived gold-standards (Figure
\ref{fig:ADNI1-xval-k-mean}). Qualitatively, both ANIMAL and ANTs-based 
segmentations demonstrate trend overlap accuracy that increases with the size of 
atlas library and template library. With 12 or more templates \mb library of at 
least 12 templates

No marked difference in segmentation accuracy is seen when either ANIMAL or ANTs
registration is used with different label fusion techniques, at any atlas or
template library sizes.  In every parameter configuration, the use of \mb with 
ANTs registration shows increased segmentation accuracy over \mb with ANIMAL registration (\todo{interaction?}).

It is interesting to note that with an even number of templates, \mb shows a
small decrease in performance relative to when one fewer template image is used
(\todo{show significance of this}). See section (\todo{ref}) for a discussion of
this behaviour. In the remainder of this section, only results from odd-sized
template libraries will be shown.

@@

ggplot(all_data_mean, 
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1)) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" )
@
  \caption{\mb mean DSC by atlas and template library size}
  \label{fig:ADNI1_xval-k_mean}
\end{figure}

With an increasing number of templates, \mb-based segmentating using ANTs
registration and majority vote label fusion shows improvement in overlap
accuracy over multi-atlas-based segmentation, using the same number of atlases
and voting method Figure \todo{ref}. The magnitude of improvement over
multi-atlas-based segmentation decreases with an increasing number of atlases,
with accuracy converging with 7 atlases.  Peak improvement in \mb accuracy (0.02
DSC) is found when one atlas is used with a template library of 20 images.
\todo{stats?}

\begin{figure}
<<ADN1-Xval-k-diff, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & method.mb=="Majority Vote" &
                templates.mb %in% seq(1,20,2)), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  coord_cartesian(ylim=c(-0.1,0.05)) + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(legend=FALSE) + 
  scale_linetype_discrete(legend=FALSE) + 
  xlab( "Number of Templates" ) + 
  ylab( "Increase in mean Kappa between MAGeT brain and multiatlas" ) 
@
  \caption{Increase in mean Kappa between \mb and multi-atlas}
  \label{ADNI1_cross_validation_MA_difference}
\end{figure}

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy
with different input atlases (Figure \todo{ref}).  Additionally, variability
decreases with the size of template library, and is significantly decreased
(p<0.05) when using 3-7 atlases and 17 or more templates.

\begin{figure}
<<ADN1-Xval-k-sd-t-test-p-plot, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ANTSmajvote = subset(all_data, reg_method=="ANTS" & method.mb=="Majority Vote")

data.stats = ddply(ANTSmajvote, c("subject", "label","atlases","templates.mb"), function (df) {
  data.frame(
    MA.sd  =  sd(df$k.ma),
    MB.sd  =  sd(df$k.mb),
    MA.var = var(df$k.ma),
    MB.var = var(df$k.mb)
  )
})

stats.tests = ddply(data.stats, c("atlases","templates.mb"), function (df) {
  t.var = t.test(df$MA.var, df$MB.var)
  t.sd  = t.test(df$MA.sd , df$MB.sd)
  data.frame(
    stat      = c("var", "sd"),
    statistic = c(t.var$statistic, t.sd$statistic),
    p.value   = c(t.var$p.value, t.sd$p.value)
  )
})

ggplot(stats.tests, 
       aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  facet_grid( . ~ stat) + 
  geom_line() + 
  geom_hline(aes(yintercept=0.05), linetype='dashed') + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + scale_y_continuous(breaks=seq(0,1,by=0.05)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" )

@
\end{figure}

<<ADNI1-Xval-sd-by-subject, cache=TRUE>>=
SDs = data.stats[c('subject','label','atlases','templates.mb','MA.sd','MB.sd')]
SDs = subset(SDs, templates.mb %in% c(5,10,15,20) & atlases == 5)

melted = melt(SDs, measure.vars=c('MA.sd','MB.sd'))
ggplot(melted, aes(x=templates.mb, y=value, 
                   group=interaction(templates.mb, variable), colour=as.factor(variable))) + 
  scale_x_continuous(breaks=c(5,10,15,20)) + 
  geom_violin() 
@

@@

\begin{figure}
<<ADN1-Xval-k-diff-ANTS-20tmpls, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
a_20_tmpls = subset(all_data_mean, 
                      reg_method  == "ANTS" &
                      method.mb   == "Majority Vote" &
                      templates.mb== 20)
ggplot(a_20_tmpls, aes(x=factor(atlases))) +
  #geom_point(aes(y=k.mb,group=1,colour="MAGeT"),position="jitter") + 
  #geom_point(aes(y=k.ma,group=2,colour="MA"),position="jitter") + 
  geom_violin(aes(y=k.mb,group=factor(atlases),colour="MAGeT")) + 
  geom_violin(aes(y=k.ma,group=factor(atlases),colour="MA")) + 
  geom_smooth(aes(y=k.mb,group=1,colour="MAGeT")) +
  geom_smooth(aes(y=k.ma,group=2,colour="MA")) +

  #stat_summary(fun.y=mean,geom='point',size=3,colour="red",aes(y=k.ma)) +
  #stat_summary(fun.y=mean,geom='line',colour="red",aes(y=k.ma,group=1)) +
  #stat_summary(fun.y=mean,geom='point',size=3,colour="blue",aes(y=k.mb)) +
  #stat_summary(fun.y=mean,geom='line',colour="blue",aes(y=k.mb,group=1)) +
  labs(x="Number of Atlases", y = "Kappa", 
       title="mean Kappa of MA and MAGeT brain (ANTS/20 templates) \nvs\natlas library size") +
  scale_colour_hue(name="Method") + 
  coord_cartesian(ylim=c(0.7,0.9))
  #theme(legend.justification=c(1,0), legend.position=c(1,0))
@
  \caption{Change in mean Kappa between multi-atlas and \mb using ANTS,
  20 templates}
  \label{}
\end{figure}
